---
title: "Getting Started with chapr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with chapr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Installation

Install the development version from GitHub:

```{r eval=FALSE}
# install.packages("remotes")
remotes::install_github("dhis2-chap/chap_r_sdk")
```

*Note: If prompted for GitHub authentication, you can skip it by pressing Enter (the repository is public).*

Load the package:

```{r setup}
library(chapr)
```

## What is the Chap R SDK?

The `chapr` package provides infrastructure for developing disease forecasting models compatible with the [Chap platform](https://github.com/dhis2/chap-core). Chap (Climate Health Analytics Platform) enables health ministries to run predictive models for disease surveillance.

This SDK simplifies model development by handling:

- **CLI creation**: Command-line interfaces for train/predict workflows
- **File I/O**: Automatic CSV loading, tsibble conversion, output formatting
- **Configuration**: YAML/JSON config parsing with schema validation
- **Validation**: Test suites to verify Chap compatibility

## Quick Start

The recommended pattern uses `create_chap_cli()` to create a complete command-line interface:

```{r eval=FALSE}
library(chapr)
library(dplyr)

# Define training function - receives loaded tsibble, not file paths
train_my_model <- function(training_data, model_configuration = list(), run_info = list()) {
  means <- training_data |>
    group_by(location) |>
    summarise(mean_cases = mean(disease_cases, na.rm = TRUE))

  return(list(means = means))
}

# Define prediction function - all inputs already loaded
predict_my_model <- function(historic_data, future_data, saved_model,
                              model_configuration = list(), run_info = list()) {
  predictions <- future_data |>
    left_join(saved_model$means, by = "location") |>
    mutate(samples = purrr::map(mean_cases, ~c(.x))) |>
    select(-mean_cases)

  return(predictions)
}

# Enable CLI with one function call
if (!interactive()) {
  create_chap_cli(train_my_model, predict_my_model)
}
```

### Command Line Usage

Save the above code as `model.R`, then use from the command line.

**In terminal:**
```bash
# Train the model
Rscript model.R train --data training_data.csv

# Generate predictions
Rscript model.R predict --historic historic.csv --future future.csv \
    --output predictions.csv

# Display model information
Rscript model.R info
```

## Model Function Interface

Your model needs two functions:

### Training Function

```{r eval=FALSE}
train_fn <- function(training_data, model_configuration = list(), run_info = list()) {
  # training_data: tsibble with time_period index, location key, disease_cases
  # model_configuration: optional list of parameters from config file
  # run_info: runtime info from Chap (prediction_length, additional_continuous_covariates, etc.)
  # Returns: model object (saved as RDS)
}
```

### Prediction Function

```{r eval=FALSE}
predict_fn <- function(historic_data, future_data, saved_model,
                       model_configuration = list(), run_info = list()) {
  # historic_data: tsibble with historical observations
  # future_data: tsibble with time periods to predict (no disease_cases)
  # saved_model: object returned by train_fn
  # run_info: runtime info from Chap
  # Returns: tibble with samples list-column
}
```

**Important**: `historic_data` may contain more recent observations than training data. Time series models should refit to `historic_data` before forecasting.

## What the SDK Handles

You don't need to write code for:

| Task | SDK handles it |
|------|----------------|
| Loading CSV files | `readr::read_csv()` |
| Converting to tsibbles | `tsibble::as_tsibble()` |
| Detecting time columns | Finds `time_period`, `date`, `week`, etc. |
| Detecting key columns | Finds `location`, `region`, etc. |
| Loading/saving models | `readRDS()` / `saveRDS()` |
| Parsing configs | `yaml::yaml.load_file()` |

Your functions only contain business logic - no file I/O boilerplate.

## Data Format

### Training/Historic Data

CSV with time, location, target, and covariates:

```
time_period,location,disease_cases,population,rainfall
2023-01,LocationA,45,10000,120.5
2023-02,LocationA,52,10000,85.2
2023-01,LocationB,78,15000,130.1
```

### Future Data

Same structure without the target variable:

```
time_period,location,population,rainfall
2023-05,LocationA,10000,95.0
2023-06,LocationA,10000,110.3
```

### Prediction Output

Tibble with `samples` list-column containing numeric vectors:

```{r eval=FALSE}
# Deterministic: single value per row
tibble(
  time_period = "2023-05",
  location = "LocationA",
  samples = list(c(42))
)

# Probabilistic: multiple Monte Carlo samples
tibble(
  time_period = "2023-05",
  location = "LocationA",
  samples = list(rpois(1000, lambda = 42))
)
```

## Configuration

### Reading Configuration Files

```{r eval=FALSE}
# Via CLI - config passed automatically to your functions
Rscript model.R train --data data.csv --config config.yaml

# Or read manually:
config <- read_model_config("config.yaml")
```

### Safe Parameter Extraction

```{r}
config <- list(
  model = list(
    params = list(learning_rate = 0.01, epochs = 100)
  )
)

# Extract nested parameters with defaults
lr <- get_config_param(config, "model", "params", "learning_rate", .default = 0.001)
print(lr)

# Returns default if path not found
missing <- get_config_param(config, "model", "missing", .default = "default")
print(missing)
```

### Configuration Schema

Define a schema for validation and the `info` subcommand:

```{r eval=FALSE}
config_schema <- create_config_schema(
  title = "My Model Configuration",
  properties = list(
    n_samples = schema_integer(default = 100L, minimum = 1L),
    learning_rate = schema_number(default = 0.01, minimum = 0, maximum = 1)
  )
)

create_chap_cli(train_fn, predict_fn, model_config_schema = config_schema)
```

## Next Steps

- **[Building Your First Chap Model](model-development-tutorial.html)**: Step-by-step tutorial with validation
- **[Working with Spatio-Temporal Data](spatio-temporal-data.html)**: Utilities for aggregation and transformation
- **[Function Reference](../reference/index.html)**: Complete API documentation

## Getting Help

- **Issues**: [GitHub Issues](https://github.com/dhis2-chap/chap_r_sdk/issues)
- **CHAP Platform**: [github.com/dhis2/chap-core](https://github.com/dhis2/chap-core)
