---
title: "Building Your First Chap Model"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Building Your First Chap Model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This tutorial walks you through building a Chap-compatible model step-by-step, using a validation-first approach to ensure your model works correctly before deploying it.

## Setup

```{r setup}
library(chapr)
library(dplyr)
```

## Step 1: Understand the Function Interface

Every Chap model requires two functions:

**Training function:**
```{r eval=FALSE}
train_fn <- function(training_data, model_configuration = list(), run_info = list()) {
  # training_data: tsibble with time_period index, location key,
  #                disease_cases, and covariates
  # model_configuration: optional list of parameters
  # run_info: runtime info from Chap (prediction_length, additional_continuous_covariates)
  # Returns: any model object (list, fitted model, etc.)
}
```

**Prediction function:**
```{r eval=FALSE}
predict_fn <- function(historic_data, future_data, saved_model,
                       model_configuration = list(), run_info = list()) {
  # historic_data: tsibble with historical observations
  # future_data: tsibble with time periods to predict
  # saved_model: the object returned by train_fn
  # run_info: runtime info from Chap
  # Returns: tibble with samples list-column containing numeric vectors
  #   - For deterministic models: single sample per row (e.g., samples = list(c(42)))
  #   - For probabilistic models: multiple samples per row (e.g., 1000 samples)
  #
  # IMPORTANT: historic_data may contain more recent data than training_data.
  # For time series models, you should refit to historic_data before forecasting.
  # Use saved_model for hyperparameters/structure, not the fitted model itself.
  # See examples/arima_model/ for a demonstration of this pattern.
}
```

## Step 2: Explore the Example Data

The SDK provides example datasets for testing. Let's examine the Laos monthly data:

```{r explore-data}
data <- get_example_data('laos', 'M')
names(data)
```

The example data contains four tsibbles. Each has `time_period` as the index and `location` as the key:

**Training data** - what your model learns from:
```{r}
data$training_data
```

**Historic data** - observations available at prediction time:
```{r}
data$historic_data
```

Notice that `historic_data` extends beyond `training_data`:

```{r}
# Training data ends at:
max(data$training_data$time_period)

# Historic data ends at:
max(data$historic_data$time_period)
```

This is a key concept: when Chap calls your prediction function, `historic_data` may contain **more recent observations** than what the model was trained on. For time series models (ARIMA, exponential smoothing, etc.), you should **refit** the model to `historic_data` before forecasting. See `examples/arima_model/` for a demonstration of this pattern.

**Future data** - time periods to predict (no `disease_cases`):
```{r}
data$future_data
```

**Example predictions** - what your model should output:
```{r}
data$predictions
```

The predictions tibble has a `samples` list-column where each element is a numeric vector. Let's look at the structure:

```{r}
# Each row has a vector of samples
data$predictions$samples[[1]]
```

For probabilistic models, each vector contains multiple Monte Carlo samples (e.g., 1000). For deterministic models, use a single sample per row: `samples = list(c(42))`.

## Step 3: Validate Before Implementing

Before writing any model logic, let's see what the validation expects. Start with stub functions:

```{r validate-stubs}
train_fn <- function(training_data, model_configuration = list(), run_info = list()) {
  list(dummy = 1)
}

predict_fn <- function(historic_data, future_data, saved_model,
                       model_configuration = list(), run_info = list()) {
  future_data
}

result <- validate_model_io(train_fn, predict_fn, data)
result$success
result$errors
```

The validation tells us exactly what's missing: the `samples` list-column in predictions.

## Step 4: Implement a Simple Mean Model

Now let's implement a minimal model that predicts the historical mean for each location.
Since all models must return a `samples` list-column, we wrap the single prediction value in a list:

```{r implement-model}
train_fn <- function(training_data, model_configuration = list(), run_info = list()) {
  means <- training_data |>
    as_tibble() |>
    summarise(mean_cases = mean(disease_cases, na.rm = TRUE), .by = location)
  list(means = means)
}

predict_fn <- function(historic_data, future_data, saved_model,
                       model_configuration = list(), run_info = list()) {
  future_data |>
    left_join(saved_model$means, by = "location") |>
    mutate(samples = purrr::map(mean_cases, ~c(.x))) |>
    select(-mean_cases)
}
```

Note: We use `as_tibble()` in the training function because `summarise(.by = ...)` needs a tibble to collapse across the time dimension. The prediction function works directly on tsibbles since `left_join()` and `mutate()` preserve tsibble structure.

## Step 5: Validate the Implementation

```{r validate-implementation}
result <- validate_model_io(train_fn, predict_fn, data)
result$success
result$n_predictions
```

The validation passes and we generated 21 predictions.

## Step 6: Validate Against All Datasets

The SDK can validate against all available example datasets:

```{r validate-all}
result <- validate_model_io_all(train_fn, predict_fn)
result$success
names(result$results)
```

## Step 7: Create the CLI

Once validation passes, wrap your model in a CLI.

First, create a new directory for your model project:

**In terminal:**
```bash
mkdir my_model
cd my_model
```

Then create a file called `model.R` inside this directory:

```{r eval=FALSE}
library(chapr)
library(dplyr)

train_fn <- function(training_data, model_configuration = list(), run_info = list()) {
  means <- training_data |>
    as_tibble() |>
    summarise(mean_cases = mean(disease_cases, na.rm = TRUE), .by = location)
  list(means = means)
}

predict_fn <- function(historic_data, future_data, saved_model,
                       model_configuration = list(), run_info = list()) {
  future_data |>
    left_join(saved_model$means, by = "location") |>
    mutate(samples = purrr::map(mean_cases, ~c(.x))) |>
    select(-mean_cases)
}

if (!interactive()) {
  create_chap_cli(train_fn, predict_fn)
}
```

## Step 8: Use the CLI

Your model is now ready for command-line use. First, export the example data to CSV files for testing.

**In R:**
```{r eval=FALSE}
# Export example data for testing
data <- get_example_data('laos', 'M')
write.csv(as.data.frame(data$training_data), "training_data.csv", row.names = FALSE)
write.csv(as.data.frame(data$historic_data), "historic.csv", row.names = FALSE)
write.csv(as.data.frame(data$future_data), "future.csv", row.names = FALSE)
```

Now you can test the CLI.

**In terminal:**
```bash
# Train the model
Rscript model.R train --data training_data.csv

# Generate predictions
Rscript model.R predict --historic historic.csv --future future.csv \
    --output predictions.csv

# Display model info
Rscript model.R info
```

The CLI automatically handles:

- Loading CSV files
- Converting to tsibbles
- Saving the model as RDS
- Writing predictions to CSV (converting nested samples to wide format)

## Step 9: Create MLproject for chap-core

To run your model with chap-core, you need an MLproject file. The SDK can generate this automatically using `generate_mlproject()`.

### Prerequisites

First, set up renv for reproducible dependencies:

```{r eval=FALSE}
renv::init()
renv::install("dhis2-chap/chap_r_sdk")
renv::snapshot()
```

### Generate the MLproject File

```{r eval=FALSE}
library(chapr)

generate_mlproject(model_name = "my_mean_model")
```

This creates an `MLproject` file like:

```yaml
name: my_mean_model
renv_env: renv.lock
entry_points:
  train:
    parameters:
      train_data: str
      model: str
    command: Rscript model.R train --data {train_data} --model {model}
  predict:
    parameters:
      historic_data: str
      future_data: str
      model: str
      out_file: str
    command: Rscript model.R predict --historic {historic_data} --future {future_data}
      --model {model} --output {out_file}
```

### Run with chap-core

Once you have the MLproject and renv.lock files, you can run your model with chap-core using either the basic `evaluate` command or the more advanced `evaluate2` command.

**Basic evaluation:**

In terminal:
```bash
chap evaluate --model-name ./my_model_directory --dataset-csv data.csv
```

**Advanced evaluation with backtesting (evaluate2):**

The `evaluate2` command provides more control over the evaluation process, including backtesting with multiple splits:

In terminal:
```bash
chap evaluate2 \
    ./my_model_directory \
    dataset.csv \
    output.nc \
    3 3 1 \
    true true '' timestamp false
```

Arguments explained:
- `./my_model_directory` - Path to your model directory containing MLproject and renv.lock
- `dataset.csv` - Input dataset for evaluation
- `output.nc` - Output file (NetCDF format) for evaluation results
- `3` - Number of periods to predict (e.g., 3 months ahead)
- `3` - Number of backtest splits
- `1` - Stride between splits
- `true` - Ignore environment setup (use system R)
- `true` - Enable debug logging
- `''` - Log file path (empty for no file logging)
- `timestamp` - Run directory type
- `false` - Whether this is a chapkit-model (false for MLproject-based models)

**Export metrics after evaluation:**

In terminal:
```bash
chap export-metrics \
    --input-files output.nc \
    --output-file metrics.csv
```

Key metrics to review:
- **RMSE**: Root Mean Square Error (lower is better)
- **MAE**: Mean Absolute Error (lower is better)
- **CRPS**: Continuous Ranked Probability Score for probabilistic forecasts (lower is better)
- **Coverage**: Should be close to the nominal level (e.g., 80% for 80% prediction intervals)

See `examples/fable_model/` for a complete example with evaluation results and `examples/arima_model/` for a simpler renv and MLproject setup.

## Probabilistic Models

For probabilistic forecasting, include multiple Monte Carlo samples instead of a single value:

```{r eval=FALSE}
predict_fn <- function(historic_data, future_data, saved_model,
                       model_configuration = list(), run_info = list()) {
  n_samples <- 1000

  future_data |>
    left_join(saved_model$means, by = "location") |>
    rowwise() |>
    mutate(
      # Generate 1000 samples from Poisson distribution
      samples = list(rpois(n_samples, lambda = mean_cases))
    ) |>
    ungroup() |>
    select(-mean_cases)
}
```

The `samples` column is a list-column where each element is a numeric vector.
The CLI automatically converts this to wide CSV format (`sample_0`, `sample_1`, ...) for Chap.

## Working with Samples

The SDK provides utility functions for working with sample-based predictions:

```{r eval=FALSE}
# Convert nested samples to wide format
wide_preds <- predictions_to_wide(nested_preds)

# Convert to long format for scoringutils
long_preds <- predictions_to_long(nested_preds)

# Compute quantiles for hub submissions
quantile_preds <- predictions_to_quantiles(nested_preds)

# Add summary statistics (mean, median, CIs)
preds_with_summary <- predictions_summary(nested_preds)
```

## Configuration Schemas

You can define a configuration schema to validate user-provided settings and provide default values. The SDK uses JSON Schema (draft-07) for validation.

### Defining a Schema

Use the schema helper functions to define type-safe configuration options:

```{r schema-definition}
my_schema <- create_config_schema(
  title = "My Model Configuration",
  description = "Configuration options for my forecasting model",
  properties = list(
    # Integer with range constraints
    n_samples = schema_integer(
      description = "Number of Monte Carlo samples for predictions",
      default = 100L,
      minimum = 1L,
      maximum = 10000L
    ),
    # Number (float) with bounds
    learning_rate = schema_number(
      description = "Learning rate for optimization",
      default = 0.01,
      minimum = 0,
      maximum = 1
    ),
    # Enum: one of a fixed set of values
    method = schema_enum(
      values = c("arima", "ets", "prophet"),
      description = "Forecasting method to use",
      default = "arima"
    ),
    # Boolean flag
    use_covariates = schema_boolean(
      description = "Whether to include covariates in the model",
      default = TRUE
    ),
    # String with optional pattern constraint
    date_format = schema_string(
      description = "Date format for output",
      default = "%Y-%m-%d"
    ),
    # Array of values
    lag_values = schema_array(
      items = list(type = "integer"),
      description = "Lag periods to include",
      default = list(1L, 2L, 3L)
    )
  ),
  required = c("n_samples")  # Mark required fields
)

# View the schema
print(my_schema)
```

### Using the Schema with CLI

Pass the schema to `create_chap_cli()` to enable automatic validation:

```{r eval=FALSE}
if (!interactive()) {
  create_chap_cli(train_fn, predict_fn, model_config_schema = my_schema)
}
```

When a user provides a configuration file (YAML or JSON), the CLI will:

1. **Validate** the config against the schema (type checking, range constraints, enum values)
2. **Apply defaults** for any missing optional parameters
3. **Report errors** with clear messages if validation fails

Example `config.yaml`:
```yaml
n_samples: 500
learning_rate: 0.05
method: ets
```

### Manual Validation

You can also validate configurations manually:

```{r manual-validation}
# Valid configuration
config <- list(n_samples = 500L, method = "ets")
result <- validate_config(config, my_schema)
result$valid

# Invalid configuration (value out of range)
bad_config <- list(n_samples = -5L)
result <- validate_config(bad_config, my_schema)
result$valid
result$errors

# Apply defaults to fill in missing values
partial_config <- list(n_samples = 200L)
full_config <- apply_config_defaults(partial_config, my_schema)
full_config$n_samples      # User value preserved
full_config$learning_rate  # Default applied
full_config$method         # Default applied
```

### Available Schema Types

| Function | Description | Key Options |
|----------|-------------|-------------|
| `schema_integer()` | Integer values | `minimum`, `maximum`, `default` |
| `schema_number()` | Numeric (float) values | `minimum`, `maximum`, `default` |
| `schema_string()` | String values | `min_length`, `max_length`, `pattern`, `default` |
| `schema_boolean()` | TRUE/FALSE values | `default` |
| `schema_enum()` | One of fixed choices | `values` (required), `default` |
| `schema_array()` | Arrays/lists | `items`, `min_items`, `max_items`, `default` |

### Adding Config Schema to MLproject

Once you have defined a configuration schema, you can include it in your MLproject file. This exposes your model's configuration options to chap-core users:

```{r eval=FALSE}
generate_mlproject(
  model_name = "my_mean_model",
  config_schema = my_schema
)
```

This adds a `user_options` section to the MLproject file:

```yaml
name: my_mean_model
renv_env: renv.lock
user_options:
  n_samples:
    type: integer
    description: Number of Monte Carlo samples for predictions
    default: 100
  learning_rate:
    type: float
    description: Learning rate for optimization
    default: 0.01
  method:
    type: str
    description: Forecasting method to use
    default: arima
entry_points:
  train:
    parameters:
      train_data: str
      model: str
    command: Rscript model.R train --data {train_data} --model {model}
  predict:
    parameters:
      historic_data: str
      future_data: str
      model: str
      out_file: str
    command: Rscript model.R predict --historic {historic_data} --future {future_data}
      --model {model} --output {out_file}
```

The `user_options` section allows chap-core to present configuration options to users and pass them to your model via a config file.

## Summary

The development workflow is:

1. **Explore** example data with `get_example_data()`
2. **Validate** with stubs using `validate_model_io()` to understand requirements
3. **Implement** your train and predict functions
4. **Validate** the implementation
5. **Test** against all datasets with `validate_model_io_all()`
6. **Deploy** with `create_chap_cli()`

## Next Steps

- See `examples/ewars_model/` for a more complex example with configuration
- See `examples/arima_model/` for a complete example with renv and MLproject integration
- Read about MLproject generation in `?generate_mlproject`
- Read about configuration schemas in `?create_config_schema`
- Explore spatial-temporal utilities in `?aggregate_temporal`
